---
title: "Study of transcriptomic response to RNAseL activation induced by 2-5A/poly I:C using bulk RNAseq"
output: html_notebook
author: Hernan Lorenzi
date: "3/11/2024"
output: html_notebook

---

\Center

*PI:* Nick Guydosh
*Point of Contact:* Agnes Karasik
*Contact email:* agnes.karasik@nih.gov

*Summary:* The goal of this project is to identify metabolic pathways and transcription factors that are turned on during the activation of the cellular response against viral infections using bulk RNAseq data.

```{r, setup, message=FALSE, error=FALSE, include=FALSE}

```

# Load libraries
```{r load libraries, message=FALSE}
suppressMessages(library("org.Hs.eg.db"))
suppressMessages(library("pheatmap"))
suppressMessages(library("EnhancedVolcano"))
suppressMessages(library("ggplot2"))
suppressMessages(library("ggpubr"))
suppressMessages(library("DESeq2"))
suppressMessages(library("stringr"))
suppressMessages(library("biomaRt"))
suppressMessages(library("tidyverse"))
suppressMessages(library("pcaExplorer"))
suppressMessages(library("VennDiagram"))
suppressMessages(library("clusterProfiler"))
suppressMessages(library("GOSemSim"))
suppressMessages(library("ggsci"))
suppressMessages(library("viridis"))
suppressMessages(library("ggrepel"))
suppressMessages(library("RColorBrewer"))
suppressMessages(library("msigdbr"))
suppressMessages(library("cowplot"))
suppressMessages(library("enrichplot"))
suppressMessages(library("ReactomePA"))
suppressMessages(library("ggupset"))
suppressMessages(library("ggraph"))
suppressMessages(library("biomaRt"))
```

# Define functions
```{r aux functions}

# Load auxyliary functions
source(file = "01_aux_rnaseq_functions.R")

# Load enrichment functions
source(file = "./02_Gene_enrichment_functions.R")

```


# Load data
```{r Loading data}

# Path to data files
RIBO_READ_COUNTS <-  "../data/Ribo_count.csv"
RNA_READ_COUNTS <- "../data/cdsrna_round.csv"

RIBO_METADATA <- "../data/Metadata_footprint.txt"
RNA_METADATA <- "../data/Metadata.txt"

# Define lists
all <- list()
metadata <- list()

all["ribo"] <- list(read.delim2(RIBO_READ_COUNTS , sep = ",", header = TRUE, row.names = 1, comment.char = c("#")))

all["rna"] <- list(read.delim2(RNA_READ_COUNTS , sep = ",", header = TRUE, row.names = 1, comment.char = c("#")))

# Make sure read counts are numeric and rounded to 0 decimals
for (idx in c("ribo","rna")){
  all.tmp <- as.data.frame(apply(all[[idx]],2, function(x){ round(as.numeric(x), digits = 0)} ))
  rownames(all.tmp) <- rownames(all[[idx]])
  all[idx] <- list(all.tmp)
  
  #Remove all zero rows
  all[[idx]] <- remove_all_zero_rows(all[[idx]] , min_total_count = 0)
}

# Load metadata
metadata["ribo"] <- list(read.delim2(RIBO_METADATA , sep = "\t", row.names = 1, header = T))
metadata["rna"] <- list(read.delim2(RNA_METADATA , sep = "\t", row.names = 1, header = T))

# Keep only metadata columns of interest
for (idx in c("ribo","rna")){
  metadata[[idx]] <- select(metadata[[idx]],c("Sample_name","Read_length","Sequencing_pool",
                            "Genotype","Group","Inducer"))
  # Sort tables so metadata and read counts match order
  meta.tmp <- metadata[[idx]]
  metadata[[idx]] <-  meta.tmp[match(colnames(all[[idx]]), rownames(meta.tmp)), ]
  
  # Add total read counts and sample id columns to metadata
  metadata[[idx]] <- cbind(metadata[[idx]], 
                       Read_counts=colSums(all[[idx]]), 
                       Sample_id = rownames(metadata[[idx]]))
  
  # Add new factors (Read_depth (high >= 10M reads / Low < 10M reads))
  meta.tmp <- metadata[[idx]]
  meta.tmp$Read_depth <- 'High'
  meta.tmp[meta.tmp$Read_counts < 1e7,]$Read_depth <- 'Low'
  meta.tmp$Read_depth <- as.factor(meta.tmp$Read_depth)
  
  meta.tmp$Batch <- 'Old'
  meta.tmp[meta.tmp$Sample_name == "n",]$Batch <- 'New'
  meta.tmp$Batch <- as.factor(meta.tmp$Batch)
  
  metadata[[idx]] <- meta.tmp

}

```

# Normalize data to TPMs to run some comparative analysis across samples
```{r}
#all.tpm <- normalize_by_TPM(counts.df = all)
```

###########################################
# Analysis of DE using DESeq2
###########################################

```{r deseq2}
for (idx in c("ribo","rna")){
  # Convert metadata to factors
  meta.tmp <- metadata[[idx]]
  for (variable in colnames(meta.tmp)[!colnames(meta.tmp) %in% "Read_counts"]){
    meta.tmp[,variable] <- as.factor(meta.tmp[,variable])  
  }
  metadata[idx] <- list(meta.tmp)
}
```

# Analysis of Dataset ONE

```{r Dataset one}

for (idx in c("ribo","rna")){
  all_one  <- all[[idx]]
  meta_one <- metadata[[idx]]
  # Generate DESeq2 object for NS and ST condition ONLY. We could potentially add Read_counts as either a covariate or as a factor (by grouping Read counts into Low and High count bins). Ideally, this should be done adding Read_counts as a random effect using mixed models, that can only be applied in Limma package. Addind it as a fixed effect in the fdesign formula might lower the power of the analysis.  
  #dds.one <- DESeqDataSetFromMatrix(countData = all_one, 
  #                              colData = meta_one,  
  #                              design = ~ Genotype + Inducer + Genotype:Inducer)
  
  # add new factors (Group_gt_ind and Read_depth (high > 10M reads / Low < 10M reads))
  meta_one$Group_gt_ind <- factor(paste0(meta_one$Genotype, meta_one$Inducer))
  meta_one$Read_depth <- 'High'
  meta_one[meta_one$Read_counts < 1e7,]$Read_depth <- 'Low'
  meta_one$Read_depth <- as.factor(meta_one$Read_depth)
  
  # Adding read_depth in design to control for read_depth
  dds.one <- DESeqDataSetFromMatrix(countData = all_one, 
                                colData = meta_one,  
                                design = ~ Read_depth + Group_gt_ind)
  
  # Plot total reads per sample using barchar
  p <- ggbarplot(data = meta_one, 
            x = "Sample_id", 
            y = "Read_counts",
            x.text.angle = 90,
            fill = "Inducer", 
            title = "Total read counts", 
            ylab = "Read counts",
            sort.by.groups = TRUE,
            palette = "jco",
            sort.val = "asc", 
            facet.by = "Genotype")
  ggsave(paste0("Plots/barplot_read_counts_",idx,".pdf"), plot = p)
  p
  
  
  # Normalize counts
  vsd.one <- vst(dds.one, blind=FALSE)
  rlog.one <- rlog(dds.one, blind=FALSE)
  
  # Keep genes with at least 10 reads total across samples
  keep <- rowSums(counts(dds.one)) >= 20
  dds.one <- dds.one[keep,]
  
  # Calculate distances between samples
  sampleDists <- dist(t(assay(vsd.one)))
  
  # Plot inter-sample distances
  old.par <- par(no.readonly=T)
  
  sampleDistMatrix <- as.matrix(sampleDists)
  rownames(sampleDistMatrix) <- paste(rlog.one$Sequencing_pool, rlog.one$Genotype, rlog.one$Inducer, sep="-")
  colnames(sampleDistMatrix) <- NULL
  colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
  pheatmap(sampleDistMatrix,
           clustering_distance_rows=sampleDists,
           clustering_distance_cols=sampleDists,
           col=colors)
  
  # PCA
  
  for (coord in c(2,3)){
    # Color by Inducer and shape by Genotype
    pcaData <- plotPCA(rlog.one, intgroup=c("Genotype", "Inducer"), 
                       returnData=TRUE, pcsToUse=c(1,coord))
    
    percentVar <- round(100 * attr(pcaData, "percentVar"))
    y.coords = c(min(pcaData$PC1, pcaData[,2]), max(pcaData$PC1, pcaData[,2]))
    x.coords = y.coords
    
    p1 <- ggplot(pcaData, aes(PC1, pcaData[,2], color=Genotype, shape=Inducer)) +
      geom_point(size=3) + 
      scale_color_lancet() + 
      xlab(paste0("PC1: ",percentVar[1],"% variance")) +
      ylab(paste0("PC",coord,": ",percentVar[2],"% variance")) +
      coord_fixed(ratio = (max(pcaData$PC1)-min(pcaData$PC1))/(max(pcaData[,2]-min(pcaData[,2])))) +
      theme_bw()
      
    ggsave2(filename = paste0("Plots/pca_dataset_Induc_gt_PC1_vs_PC",coord,"_",idx,".pdf"), plot = p1)
    
    p1
    
    # Color by Read_counts and shape by Inducer
    pcaData <- plotPCA(rlog.one, intgroup=c("Read_counts", "Inducer"), 
                       returnData=TRUE, pcsToUse=c(1,coord))
    percentVar <- round(100 * attr(pcaData, "percentVar"))
    
    p2 <- ggplot(pcaData, aes(PC1, pcaData[,2], color=Read_counts, shape=Inducer)) +
      geom_point(size=3) +
      xlab(paste0("PC1: ",percentVar[1],"% variance")) +
      ylab(paste0("PC",coord,": ",percentVar[2],"% variance")) + 
      coord_fixed(ratio = (max(pcaData$PC1)-min(pcaData$PC1))/(max(pcaData[,2]-min(pcaData[,2])))) + scale_color_gradient2(high = "blue", mid = "yellow") +
      theme_bw()
    ggsave(paste0("Plots/pca_dataset_1_Induc_read_counts_PC1_vs_PC",coord,"_",idx,".pdf"), plot = p2)
    p2
    
    # Color by Read_depth and shape by Inducer
    pcaData <- plotPCA(rlog.one, intgroup=c("Read_depth", "Inducer"), 
                       returnData=TRUE, pcsToUse=c(1,coord))
    percentVar <- round(100 * attr(pcaData, "percentVar"))
    p3 <- ggplot(pcaData, aes(PC1, pcaData[,2], color=Read_depth, shape=Inducer)) +
      geom_point(size=3) + scale_color_lancet() + 
      xlab(paste0("PC1: ",percentVar[1],"% variance")) +
      ylab(paste0("PC",coord,": ",percentVar[2],"% variance")) + 
      coord_fixed(ratio = (max(pcaData$PC1)-min(pcaData$PC1))/(max(pcaData[,2])-min(pcaData[,2]))) +
      theme_bw()
    
    ggsave(paste0("Plots/pca_dataset_1_Depth_Induc_PC1_vs_PC",coord,"_",idx,".pdf"), plot = p3)
    p3
    
    # Color by Batch and shape by Inducer
    pcaData <- plotPCA(rlog.one, intgroup=c("Batch", "Inducer"), 
                       returnData=TRUE, pcsToUse=c(1,coord))
    percentVar <- round(100 * attr(pcaData, "percentVar"))
    p4 <- ggplot(pcaData, aes(PC1, pcaData[,2], color=Batch, shape=Inducer)) +
      geom_point(size=3) + scale_color_lancet() + 
      xlab(paste0("PC1: ",percentVar[1],"% variance")) +
      ylab(paste0("PC",coord,": ",percentVar[2],"% variance")) + 
      coord_fixed(ratio = (max(pcaData$PC1)-min(pcaData$PC1))/(max(pcaData[,2])-min(pcaData[,2]))) +
      theme_bw()
    
    ggsave(paste0("Plots/pca_dataset_1_Batch_Induc_PC1_vs_PC",coord,"_",idx,".pdf"), plot = p4)
    p4
   
  }
}
```

### resultsNames(dds)

# Filtering out poorly-expressed genes (less than 20 reads across all samples) 
```{r filtering genes based on total counts of reads}

# Keep genes with at least 20 reads total across samples
keep <- rowSums(counts(dds.one) >= 20 ) >= 3
dds.one <- dds.one[keep,]

  
dds.one.wt <- dds.one[ , dds.one$Genotype == "WT"]
dds.one.wt$Genotype <- droplevels(dds.one.wt$Genotype)
dds.one.wt$Group_gt_ind <- droplevels( dds.one.wt$Group_gt_ind)
dds.one.wt$Group <- droplevels( dds.one.wt$Group)

dds.one.ko <- dds.one[ , dds.one$Genotype == "RNaseL_KO"]
dds.one.ko$Genotype <- droplevels( dds.one.ko$Genotype)
dds.one.ko$Group_gt_ind <- droplevels( dds.one.ko$Group_gt_ind)
dds.one.ko$Group <- droplevels( dds.one.ko$Group)

```



# Using groups instead of interactions

```{r}

# Fetch gene symbols from  Biomart
  ensembl_list <- rownames(all_one)
    
  # 1- fetch human annotation
  ensembl_h <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
  
  # 2- Search symbols in hgnc_symbol column
  ensembl_to_symbol <- getBM(attributes = c("ensembl_gene_id",  "hgnc_symbol"),
                    filters = "ensembl_gene_id",
                    values = ensembl_list,
                    mart = ensembl_h)
  # 3- Rename columns
  colnames(ensembl_to_symbol) <- c("Ensembl_ID", "gene_name")

# Make output directory for DE analysis
dir.create(path = "./DE", showWarnings = F)

# Calculate DE for WT samples
design(dds.one.wt) <- ~ Batch + Group_gt_ind
dds.one.wt$Group_gt_ind <- relevel(dds.one.wt$Group_gt_ind, "WTNone")
dds.one.wt <- DESeq(dds.one.wt)
resultsNames(dds.one.wt)

# design(dds.one.wt) <- ~ Group_gt_ind
# dds.one.wt$Group_gt_ind <- relevel(dds.one.wt$Group_gt_ind, "WTNone")
# dds.one.wt <- DESeq(dds.one.wt)
# resultsNames(dds.one.wt)
# 
# design(dds.one.wt) <- ~ Read_depth + Group_gt_ind
# dds.one.wt$Group_gt_ind <- relevel(dds.one.wt$Group_gt_ind, "WTNone")
# dds.one.wt <- DESeq(dds.one.wt)
# resultsNames(dds.one.wt)

# Using lfcShrink instead of results to reduce high Log2FC bias of genes with low expression
res_wtIC_vs_wtNone <- lfcShrink(dds.one.wt, coef = "Group_gt_ind_WTpolyIC_vs_WTNone", type = "ashr", )
res_wt25A_vs_wtNone <- lfcShrink(dds.one.wt, coef = "Group_gt_ind_WT25A_vs_WTNone", type = "ashr")

summary(res_wtIC_vs_wtNone, alpha = 0.05)
summary(res_wt25A_vs_wtNone, alpha = 0.05)

# Calculate DE for KO samples
dds.one.ko$Group_gt_ind <- relevel(dds.one.ko$Group_gt_ind, "RNaseL_KONone")
dds.one.ko <- DESeq(dds.one.ko)
resultsNames(dds.one.ko)

res_koIC_vs_koNone <- lfcShrink(dds.one.ko, coef = "Group_gt_ind_RNaseL_KOpolyIC_vs_RNaseL_KONone", type = "ashr")
res_ko25A_vs_koNone <- lfcShrink(dds.one.ko, coef = "Group_gt_ind_RNaseL_KO25A_vs_RNaseL_KONone", type = "ashr")

summary(res_koIC_vs_koNone, alpha = 0.05)
summary(res_ko25A_vs_koNone, alpha = 0.05)

# Define function for processing and saving result tables
sort_and_write_res_table <- function(result_table, file_name){
  # Sort genes by (padj)
  result_table_sorted <- result_table[order(result_table$padj, decreasing = FALSE),]
  # Add gene symbols
  gene_list <- rownames(result_table_sorted)
  symbol_list <- ensembl_to_symbol$gene_name[match(gene_list, ensembl_to_symbol$Ensembl_ID)]
  df <-as.data.frame(cbind(result_table_sorted, Gene_name = symbol_list))
  
  # Write sorted table to file
  write.table(df, file = paste0("./DE/",file_name,".txt"), 
            sep = "\t", col.names=NA)
  return(result_table_sorted)
}

# Sort results by Log2FC
res_wtIC_vs_wtNone.logfc_sorted <- sort_and_write_res_table(res_wtIC_vs_wtNone, "DE_wtIC_vs_wtNone")
res_koIC_vs_koNone.logfc_sorted <- sort_and_write_res_table(res_koIC_vs_koNone, "DE_koIC_vs_koNone")
res_wt25A_vs_wtNone.logfc_sorted <- sort_and_write_res_table(res_wt25A_vs_wtNone, "DE_wt25A_vs_wtNone")
res_ko25A_vs_koNone.logfc_sorted <- sort_and_write_res_table(res_ko25A_vs_koNone, "DE_ko25A_vs_koNone")

# Save sorted files as a list
DE_results = list()
DE_results[["wtIC_vs_wtNone"]]  <- res_wtIC_vs_wtNone.logfc_sorted
DE_results[["koIC_vs_koNone"]] <- res_koIC_vs_koNone.logfc_sorted
DE_results[["wt25A_vs_wtNone"]] <- res_wt25A_vs_wtNone.logfc_sorted
DE_results[["ko25A_vs_koNone"]] <- res_ko25A_vs_koNone.logfc_sorted

# Generate result table of DE genes in  res_wtIC_vs_wtNone but not in res_koIC_vs_koNone
# or in res_wt25A_vs_wtNone. 
# This are genes that might require both an induction via dsRNA/DNA and a functional RNAseL


# Filter 1 adj.pvalues > 0.25 in other samples
res_wtIC_vs_wtNone.sig.id <- rownames(res_wtIC_vs_wtNone)[res_wtIC_vs_wtNone$padj <= 0.05 & !is.na(res_wtIC_vs_wtNone$padj)]
res_koIC_vs_koNone.sig.id <- rownames(res_koIC_vs_koNone)[res_koIC_vs_koNone$padj <= 0.25]
res_wt25A_vs_wtNone.sig.id <- rownames(res_wt25A_vs_wtNone)[res_wt25A_vs_wtNone$padj <= 0.25]

filtered_ids <- res_wtIC_vs_wtNone.sig.id[!(res_wtIC_vs_wtNone.sig.id %in% c(res_wt25A_vs_wtNone.sig.id, res_koIC_vs_koNone.sig.id))]

res_wtIC_vs_wtNone.filtered <- subset(res_wtIC_vs_wtNone, rownames(res_wtIC_vs_wtNone) %in% filtered_ids)
res_koIC_vs_koNone.filtered <- subset(res_koIC_vs_koNone, rownames(res_koIC_vs_koNone) %in% filtered_ids)
res_wt25A_vs_wtNone.filtered <- subset(res_wt25A_vs_wtNone, rownames(res_wt25A_vs_wtNone) %in% filtered_ids)

# Filter 2: abs(Log2FC) res_wt25A_vs_wtNone.filtered < 0.1 
#filter2 <- log2(abs(res_wtIC_vs_wtNone.filtered$log2FoldChange)/abs(res_wt25A_vs_wtNone.filtered$log2FoldChange)) > log2(3/1) #Log2FC difference >= 3
filter2 <- abs(res_wt25A_vs_wtNone.filtered$log2FoldChange) < 0.1

# Filter 3: opposite Log2FC
filter3 <- res_wtIC_vs_wtNone.filtered$log2FoldChange * res_wt25A_vs_wtNone.filtered$log2FoldChange < 0

# Apply filter2 and 3 to res_wtIC_vs_wtNone.filtered
res_wtIC_vs_wtNone.filtered23 <-  res_wtIC_vs_wtNone.filtered[filter2 | filter3, ]



# Save res_wtIC_vs_wtNone.filtered object in list
DE_results[["wtIC_vs_wtNone_KOns_25Ans"]] <- res_wtIC_vs_wtNone.filtered
DE_results[["wtIC_vs_wtNone_KOns_25Ans_f23"]] <- res_wtIC_vs_wtNone.filtered23
res_wtIC_vs_wtNone.filtered.logfc_sorted <- sort_and_write_res_table(res_wtIC_vs_wtNone.filtered, "wtIC_vs_wtNone_KOns_25Ans")
res_wtIC_vs_wtNone.filtered.logfc_sorted <- sort_and_write_res_table(res_wtIC_vs_wtNone.filtered23, "wtIC_vs_wtNone_KOns_25Ans_f23")

print("WT-polyIC vs WT-None")
summary(res_wtIC_vs_wtNone.logfc_sorted, alpha = 0.05)
print("KO-polyIC vs KO-None")
summary(res_koIC_vs_koNone.logfc_sorted, alpha = 0.05)
print("WT-25A vs WT-None")
summary(res_wt25A_vs_wtNone.logfc_sorted, alpha = 0.05)
print("KO-25A vs KO-None")
summary(res_ko25A_vs_koNone.logfc_sorted, alpha = 0.05)

# Genes significantly DE in WT and induced by polyIC
wt.g <- rownames(res_wtIC_vs_wtNone.logfc_sorted[!is.na(res_wtIC_vs_wtNone.logfc_sorted$padj) & res_wtIC_vs_wtNone.logfc_sorted$padj <= 0.05,])
# Genes significantly DE in KO and induced by polyIC
ko.g <- rownames(res_koIC_vs_koNone.logfc_sorted[!is.na(res_koIC_vs_koNone.logfc_sorted$padj) & res_koIC_vs_koNone.logfc_sorted$padj <= 0.05,])
# Genes significantly DE in WT and induced by 25A
wt.25A.g <- rownames(res_wt25A_vs_wtNone.logfc_sorted[!is.na(res_wt25A_vs_wtNone.logfc_sorted$padj) & res_wt25A_vs_wtNone.logfc_sorted$padj <= 0.05,])
# Genes significantly DE in KO and induced by 25A
ko.25A.g <- rownames(res_ko25A_vs_koNone.logfc_sorted[!is.na(res_ko25A_vs_koNone.logfc_sorted$padj) & res_ko25A_vs_koNone.logfc_sorted$padj <= 0.05,])
```


# DE analysis between pIC WT and 25A WT
```{r}
# Calculate DE between pIC WT and 25A WT samples
dds.one.wt$Group_gt_ind <- relevel(dds.one.wt$Group_gt_ind, "WT25A")
dds.one.wt <- DESeq(dds.one.wt)
resultsNames(dds.one.wt)


# Using lfcShrink instead of results to reduce high Log2FC bias of genes with low expression
res_wtIC_vs_wt25A <- lfcShrink(dds.one.wt, coef = "Group_gt_ind_WTpolyIC_vs_WT25A", type = "ashr", )

summary(res_wtIC_vs_wt25A, alpha = 0.05)

# Save results
res_wtIC_vs_wt25A.logfc_sorted <- sort_and_write_res_table(res_wtIC_vs_wt25A, "DE_wtIC_vs_wt25A")

gene_ids_wt25A_vs_wtNone_lowFC <- rownames(subset(res_wt25A_vs_wtNone, res_wt25A_vs_wtNone$log2FoldChange <= 0.1))
res_wtIC_vs_wt25A_sig_fc1 <- subset(res_wtIC_vs_wt25A, res_wtIC_vs_wt25A$padj <= 0.05 & rownames(res_wtIC_vs_wt25A) %in% gene_ids_wt25A_vs_wtNone_lowFC )
res_wtIC_vs_wt25A_sig_fc1.logfc_sorted <- sort_and_write_res_table(res_wtIC_vs_wt25A_sig_fc1, "DE_wtIC_vs_wt25A_sig_fc1")

# Save res_wtIC_vs_wt25A and res_wtIC_vs_wtNone.filtered objects in DE_results list
DE_results[["wtIC_vs_wt25A"]] <- res_wtIC_vs_wt25A.logfc_sorted
DE_results[["wtIC_vs_wt25A_sig_f1"]] <- res_wtIC_vs_wt25A_sig_fc1.logfc_sorted

```


# Plot Venn Diagram with DE gene overlaps among treatments.
```{r}
library("ggvenn")

ggvenn(
  list("WT_plyI:C"=wt.g, "KO_polyI:C"=ko.g, "WT_25A"=wt.25A.g ), 
  fill_color = c("#440154ff", "#21908dff", "#fde725ff"),
  stroke_size = 0.5, set_name_size = 4
  )
```

```{r heatmap}
# Select factors to be displayed together with the expression data
df.ns <- as.data.frame(colData(dds.one)[,c("Genotype","Inducer")])

# plot_heat_map
plot_heat_map(my_vstd = vsd.one, gene_list = unique(c(wt.g[1:30],wt.25A.g[1:30],ko.g[1:30])), file_name = "Plots/top10_per_treatment", variables = df.ns)

# Genes significantly DE in WT and not KO when induced by polyIC
wt_no_ko.g <- wt.g[!wt.g %in% ko.g]
# Genes significantly DE in KO and not WT when induced by polyIC
ko_no_wt.g <- ko.g[!ko.g %in% wt.g]
# Genes significantly DE in WT and not KO when induced by 25A
wt_no_ko.25A.g <- wt.25A.g[!wt.25A.g %in% ko.25A.g]
# Genes significantly DE in KO and not WT when induced by 25A
ko_no_wt.25A.g <- ko.25A.g[!ko.25A.g %in% wt.25A.g]


# Results df with genes DE in WT but not in KO
res_wtIC_vs_wtNone_no_ko <- res_wtIC_vs_wtNone.logfc_sorted[wt_no_ko.g, ]
res_wt25A_vs_wtNone_no_ko <- res_wt25A_vs_wtNone.logfc_sorted[wt_no_ko.25A.g,]
x <- sort_and_write_res_table(res_wtIC_vs_wtNone_no_ko, file_name = "res_wtIC_vs_wtNone_no_ko")
x <- sort_and_write_res_table(res_wt25A_vs_wtNone_no_ko, file_name = "res_wt25A_vs_wtNone_no_ko")
```

Plot normalized (vst) gene expression per sample
```{r Plot normalized (vst) gene expression per sample}

# Plot genes with most up and down regulation with each inducer
genes_of_interest = unique(c(wt.g[1:30],wt.25A.g[1:30],ko.g[1:30]))

plot_normalized_gene_expression(my_dds = dds.one, 
                                ensmbl_gene_list = genes_of_interest, 
                                file_prefix = "Genes_with_extreme_up_down_expr_DS_ONE")

```

```{r volcano plots}
genes_of_interest_symbols <- c("RSAD2", "IFNL1", "IFNL2", "IFNB1", "RAET1L","OAS2","EGR1","DHX58","OASL","IFIH1","IFIT1","IFIT2","IFIT3","IL6","NFKBIZ","DDX58","ZC3HAV1","CXCL2","CXCL8","STAT1","MYC","SRRM2","ND4","GAPDH","ACTB","RPL7","RNY3","RNY4","EIF2A","POLR3K")


# Print volcano plots for each condition
generate_volcano_plot_with_ids(res.tmp = res_wtIC_vs_wtNone, my_file_name = "res_wtIC_vs_wtNone.logfc_sorted", gene_list = genes_of_interest_symbols)
generate_volcano_plot_with_ids(res_koIC_vs_koNone, my_file_name = "res_koIC_vs_koNone.logfc_sorted", gene_list = genes_of_interest_symbols)
generate_volcano_plot_with_ids(res_wt25A_vs_wtNone, my_file_name = "res_wt25A_vs_wtNone.logfc_sorted", gene_list = genes_of_interest_symbols)
generate_volcano_plot_with_ids(res_ko25A_vs_koNone, my_file_name = "res_ko25A_vs_koNone.logfc_sorted", gene_list = genes_of_interest_symbols)

# Print AGNES' style volcano plots for each condition for PAPER

agnes_volcanoplot(res.tmp = res_wtIC_vs_wtNone, my_file_name = "res_wtIC_vs_wtNone.logfc_sorted", gene_list = genes_of_interest_symbols, my_comparison = 'poly I:C vs Control')
agnes_volcanoplot(res.tmp = res_koIC_vs_koNone, my_file_name = "res_koIC_vs_koNone.logfc_sorted", gene_list = genes_of_interest_symbols, my_comparison = 'poly I:C vs Control')
agnes_volcanoplot(res.tmp = res_wt25A_vs_wtNone, my_file_name = "res_wt25A_vs_wtNone.logfc_sorted", gene_list = genes_of_interest_symbols, my_comparison = '2-5A vs Control')
agnes_volcanoplot(res.tmp = res_ko25A_vs_koNone, my_file_name = "res_ko25A_vs_koNone.logfc_sorted", gene_list = genes_of_interest_symbols, my_comparison = '2-5A vs Control')

```

```{r Burke et al Fig 4B pIC WT vs None WT}

all_one.tpm <- all_one

# Sort gene symbols based on DE results df
gene_ids <- rownames(res_wtIC_vs_wtNone)
matched.tpms <- all_one.tpm[match(gene_ids, table = rownames(all_one.tpm)),]
matched.symbols <- ensembl_to_symbol[match(gene_ids, table = ensembl_to_symbol$Ensembl_ID),]

# Build input df for Burke's plot
df <- as.data.frame(cbind(
                      Log2_pICWT_pICKO = res_wtIC_vs_wtNone$log2FoldChange -
                        res_koIC_vs_koNone$log2FoldChange, 
                      Gene_symbols = matched.symbols$gene_name,
                      tpm_pICWT = log10(rowMeans(
                        matched.tpms[,meta_one$Group_gt_ind == "WTpolyIC"])),
                      tpm_NoneWT = log10(rowMeans(
                        matched.tpms[,meta_one$Group_gt_ind == "WTNone"]))
                      )
                    )

# Remove NAs from df
df <- df[!is.na(df$tpm_NoneWT), ]

# Convert relevant char cols to numeric
df$Log2_pICWT_pICKO <- as.numeric(df$Log2_pICWT_pICKO) 
df$tpm_pICWT <- as.numeric(df$tpm_pICWT)
df$tpm_NoneWT <- as.numeric(df$tpm_NoneWT)

# Add fake gene entries to standardize gradient colors
df <- as.data.frame(rbind(df, "NA.1" =c(5,NA,NA,NA), "NA.2" =c(-5,NA,NA,NA)) )

# Change -Inf values due to Log10(0) to 0
#df$tpm_pICWT <- ifelse(df$tpm_pICWT == -Inf, 0, df$tpm_pICWT)
#df$tpm_NoneWT <- ifelse(df$tpm_NoneWT == -Inf, 0, df$tpm_NoneWT)

# Set extreme differences between Log2FCs to either -5 or 5 for coloring purposes
df[df$Log2_pICWT_pICKO <= -5, ]$Log2_pICWT_pICKO <- -5
df[df$Log2_pICWT_pICKO >= 5, ]$Log2_pICWT_pICKO <- 5

# Generate first plot with all genes
my_gradient <- c("red4","red3" ,"red2" ,"red1", "lightgray","royalblue1" ,"royalblue2","royalblue3","royalblue4")
my_title <- "Differential expression plot of mRNAs in WT cells treated or not with pIC"
p3 <- ggplot(df, aes(x=tpm_NoneWT, y=tpm_pICWT, label = Gene_symbols)) + 
  labs(title = my_title) +
  xlab(bquote('WT-None '* ~Log[10]*'(TPM)')) +
  ylab(bquote('WT-pI:C '* ~Log[10]*'(TPM)')) +
  geom_point(aes(colour=Log2_pICWT_pICKO), size = 0.5, alpha = 1 ) + 
  xlim(-3,6) + ylim(-3,6) +
  geom_abline(slope = 1, intercept = 0, col = "black", size=0.5, linetype="dashed") +
  theme_minimal() + gradient_color(my_gradient) + 
  theme(
    legend.direction = "horizontal",  
    legend.position = c(.95, .10),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    legend.margin = margin(3, 3, 3, 3),
    legend.text = element_text(size = 8),
    legend.title = element_text(face = "bold" ,size = 8, vjust = 0.9))

# Change legend title
p3$labels$colour = paste("log2(pI:C/None)WT - \nlog2(pI:C/None)KO" )

# Making genes of interest bigger
#genes_of_interesst <- ifelse(df$Log2_pICWT_pICKO > 4 | df$Log2_pICWT_pICKO < -2,  as.character(df$Gene_symbols),NA)
my_list <- c("RSAD2", "IFNL1", "IFNL2", "IFNB1", "RAET1L","OAS2","EGR1","DHX58","OASL","IFIH1","IFIT1","IFIT2","IFIT3","IL6","NFKBIZ","DDX58","ZC3HAV1","CXCL2","CXCL8","STAT1","MYC","SRRM2","ND4","GAPDH","ACTB","RPL7","RNY3","RNY4","EIF2A","POLR3K")
genes_of_interest <- ifelse(df$Gene_symbols %in% my_list,  as.character(df$Gene_symbols),NA)

df.subset <- subset(df, Gene_symbols %in% my_list)
df.subset <- as.data.frame(rbind(df.subset, "NA.1" =c(5,NA,NA,NA), "NA.2" =c(-5,NA,NA,NA)) )
p3 <- p3 + geom_point(data = df.subset, aes(x = tpm_NoneWT, y = tpm_pICWT, fill=Log2_pICWT_pICKO), size = 3, pch=21, colour="black", show.legend = FALSE) + gradient_fill(my_gradient)

# Adding labels to genes of interest
p3 <- p3 + geom_text_repel(aes(label = genes_of_interest),
                  colours = "red",
                  box.padding   = 0.4, 
                  point.padding = 0.3,
                  segment.color = 'green',
                  na.rm = TRUE,
                  size = 3,
                  min.segment.length = 0.02,
                  direction = "both",
                  segment.curvature = -0.1,
                  segment.ncp = 3,
                  segment.angle = 20,
                  max.iter = 1e5)

# Saving final plot
ggsave(filename = "burke4B.pdf", plot = p3, path = "Plots/" )

p3
```

```{r Burke et al Fig 4B 25A}

# Sort gene symbols based on DE results df
gene_ids <- rownames(res_wt25A_vs_wtNone)
matched.tpms <- all_one.tpm[match(gene_ids, table = rownames(all_one.tpm)),]
matched.symbols <- ensembl_to_symbol[match(gene_ids, table = ensembl_to_symbol$Ensembl_ID),]

# Build input df for Burke's plot
df <- as.data.frame(cbind(
                      Log2_25AWT_pICWT = res_wt25A_vs_wtNone$log2FoldChange - res_wtIC_vs_wtNone$log2FoldChange, 
                      Gene_symbols = matched.symbols$gene_name,
                      tpm_25AWT = log10(rowMeans(
                        matched.tpms[,meta_one$Group_gt_ind == "WT25A"])),
                      tpm_NoneWT = log10(rowMeans(
                        matched.tpms[,meta_one$Group_gt_ind == "WTNone"]))
                      )
                    )

# Remove NAs from df
df <- df[!is.na(df$tpm_NoneWT), ]

# Convert relevant char cols to numeric
df$Log2_25AWT_pICWT <- as.numeric(df$Log2_25AWT_pICWT) 
df$tpm_25AWT <- as.numeric(df$tpm_25AWT)
df$tpm_NoneWT <- as.numeric(df$tpm_NoneWT)

# Add fake gene entries to standardize gradient colors
df <- as.data.frame(rbind(df, "NA.1" =c(5,NA,NA,NA), "NA.2" =c(-5,NA,NA,NA)) )

# Change -Inf values due to Log10(0) to 0
#df$tpm_pICWT <- ifelse(df$tpm_pICWT == -Inf, 0, df$tpm_pICWT)
#df$tpm_NoneWT <- ifelse(df$tpm_NoneWT == -Inf, 0, df$tpm_NoneWT)

# Set extreme differences between Log2FCs to either -5 or 5 for coloring purposes
df[df$Log2_25AWT_pICWT <= -5, ]$Log2_25AWT_pICWT <- -5
df[df$Log2_25AWT_pICWT >= 5, ]$Log2_25AWT_pICWT <- 5

# Generate first plot with all genes
my_gradient <- c("red4","red3" ,"red2" ,"red1", "lightgray","royalblue1" ,"royalblue2","royalblue3","royalblue4")
my_title <- "Differential expression plot of mRNAs in WT cells treated or not with 25A"
p4 <- ggplot(df, aes(x=tpm_NoneWT, y=tpm_25AWT, label = Gene_symbols)) + 
  labs(title = my_title) +
  xlab(bquote('WT-None '* ~Log[10]*'(TPM)')) +
  ylab(bquote('WT-25A '* ~Log[10]*'(TPM)')) +
  geom_point(aes(colour=Log2_25AWT_pICWT), size = 0.5, alpha = 1 ) + 
  xlim(-3,6) + ylim(-3,6) +
  geom_abline(slope = 1, intercept = 0, col = "black", size=0.5, linetype="dashed") +
  theme_minimal() + gradient_color(my_gradient) + 
  theme(
    legend.direction = "horizontal",  
    legend.position = c(.95, .10),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    legend.margin = margin(3, 3, 3, 3),
    legend.text = element_text(size = 8),
    legend.title = element_text(face = "bold" ,size = 8, vjust = 0.9))

# Change legend title
p4$labels$colour = paste("log2(25A/None)WT - \nlog2(pI:C/None)WT" )

# Making genes of interest bigger
#genes_of_interesst <- ifelse(df$Log2_pICWT_pICKO > 4 | df$Log2_pICWT_pICKO < -2,  as.character(df$Gene_symbols),NA)
my_list <- c("RSAD2", "IFNL1", "IFNL2", "IFNB1", "RAETL1","OAS2","EGR1","DHX58","OASL","IFIH1","IFIT1","IFIT2","IFIT3","IL6","NFKBIZ","DDX58","ZC3HAV1","CXCL2","CXCL8","STAT1","MYC","SRRM2","ND4","GAPDH","ACTB","RPL7","RNY3","RNY4","EIF2A","POLR3K")
genes_of_interest <- ifelse(df$Gene_symbols %in% my_list,  as.character(df$Gene_symbols),NA)

df.subset <- subset(df, Gene_symbols %in% my_list)
df.subset <- as.data.frame(rbind(df.subset, "NA.1" =c(5,NA,NA,NA), "NA.2" =c(-5,NA,NA,NA)) )
p4 <- p4 + geom_point(data = df.subset, aes(x = tpm_NoneWT, y = tpm_25AWT, fill=Log2_25AWT_pICWT), size = 3, pch=21, colour="black", show.legend = FALSE) + gradient_fill(my_gradient)

# Adding labels to genes of interest
p4 <- p4 + geom_text_repel(aes(label = genes_of_interest),
                  colours = "red",
                  box.padding   = 0.4, 
                  point.padding = 0.3,
                  segment.color = 'green',
                  na.rm = TRUE,
                  size = 3,
                  min.segment.length = 0.02,
                  direction = "both",
                  segment.curvature = -0.1,
                  segment.ncp = 3,
                  segment.angle = 20,
                  max.iter = 1e5)

# Saving final plot
ggsave(filename = "burke4B_25Awt_pICwt.pdf", plot = p4, path = "Plots/" )

p4
```

```{r Burke et al Fig 4B 25A WT vs KO}

# Sort gene symbols based on DE results df
gene_ids <- rownames(res_wt25A_vs_wtNone)
matched.tpms <- all_one.tpm[match(gene_ids, table = rownames(all_one.tpm)),]
matched.symbols <- ensembl_to_symbol[match(gene_ids, table = ensembl_to_symbol$Ensembl_ID),]

# Build input df for Burke's plot
df <- as.data.frame(cbind(
                      Log2_25AWT_25AKO = res_wt25A_vs_wtNone$log2FoldChange - res_ko25A_vs_koNone$log2FoldChange, 
                      Gene_symbols = matched.symbols$gene_name,
                      tpm_25AWT = log10(rowMeans(
                        matched.tpms[,meta_one$Group_gt_ind == "WT25A"])),
                      tpm_NoneWT = log10(rowMeans(
                        matched.tpms[,meta_one$Group_gt_ind == "WTNone"]))
                      )
                    )

# Remove NAs from df
df <- df[!is.na(df$tpm_NoneWT), ]

# Convert relevant char cols to numeric
df$Log2_25AWT_25AKO <- as.numeric(df$Log2_25AWT_25AKO) 
df$tpm_25AWT <- as.numeric(df$tpm_25AWT)
df$tpm_NoneWT <- as.numeric(df$tpm_NoneWT)

# Add fake gene entries to standardize gradient colors
df <- as.data.frame(rbind(df, "NA.1" =c(5,NA,NA,NA), "NA.2" =c(-5,NA,NA,NA)) )

# Change -Inf values due to Log10(0) to 0
#df$tpm_pICWT <- ifelse(df$tpm_pICWT == -Inf, 0, df$tpm_pICWT)
#df$tpm_NoneWT <- ifelse(df$tpm_NoneWT == -Inf, 0, df$tpm_NoneWT)

# Set extreme differences between Log2FCs to either -5 or 5 for coloring purposes
df[df$Log2_25AWT_25AKO <= -5, ]$Log2_25AWT_25AKO <- -5
df[df$Log2_25AWT_25AKO >= 5, ]$Log2_25AWT_25AKO <- 5

# Generate first plot with all genes
my_gradient <- c("red4","red3" ,"red2" ,"red1", "lightgray","royalblue1" ,"royalblue2","royalblue3","royalblue4")
my_title <- "Differential expression plot of mRNAs in WT cells treated or not with 25A"
p5 <- ggplot(df, aes(x=tpm_NoneWT, y=tpm_25AWT, label = Gene_symbols)) + 
  labs(title = my_title) +
  xlab(bquote('WT-None '* ~Log[10]*'(TPM)')) +
  ylab(bquote('WT-25A '* ~Log[10]*'(TPM)')) +
  geom_point(aes(colour=Log2_25AWT_25AKO), size = 0.5, alpha = 1 ) + 
  xlim(-3,6) + ylim(-3,6) +
  geom_abline(slope = 1, intercept = 0, col = "black", size=0.5, linetype="dashed") +
  theme_minimal() + gradient_color(my_gradient) + 
  theme(
    legend.direction = "horizontal",  
    legend.position = c(.95, .10),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    legend.margin = margin(3, 3, 3, 3),
    legend.text = element_text(size = 8),
    legend.title = element_text(face = "bold" ,size = 8, vjust = 0.9))

# Change legend title
p5$labels$colour = paste("log2(25A/None)WT - \nlog2(25A/None)KO" )

# Making genes of interest bigger
#genes_of_interesst <- ifelse(df$Log2_pICWT_pICKO > 4 | df$Log2_pICWT_pICKO < -2,  as.character(df$Gene_symbols),NA)
my_list <- c("RSAD2", "IFNL1", "IFNL2", "IFNB1", "RAETL1","OAS2","EGR1","DHX58","OASL","IFIH1","IFIT1","IFIT2","IFIT3","IL6","NFKBIZ","DDX58","ZC3HAV1","CXCL2","CXCL8","STAT1","MYC","SRRM2","ND4","GAPDH","ACTB","RPL7","RNY3","RNY4","EIF2A","POLR3K")
genes_of_interest <- ifelse(df$Gene_symbols %in% my_list,  as.character(df$Gene_symbols),NA)

df.subset <- subset(df, Gene_symbols %in% my_list)
df.subset <- as.data.frame(rbind(df.subset, "NA.1" =c(5,NA,NA,NA), "NA.2" =c(-5,NA,NA,NA)) )
p5 <- p5 + geom_point(data = df.subset, aes(x = tpm_NoneWT, y = tpm_25AWT, fill=Log2_25AWT_25AKO), size = 3, pch=21, colour="black", show.legend = FALSE) + gradient_fill(my_gradient)

# Adding labels to genes of interest
p5 <- p5 + geom_text_repel(aes(label = genes_of_interest),
                  colours = "red",
                  box.padding   = 0.4, 
                  point.padding = 0.3,
                  segment.color = 'green',
                  na.rm = TRUE,
                  size = 3,
                  min.segment.length = 0.02,
                  direction = "both",
                  segment.curvature = -0.1,
                  segment.ncp = 3,
                  segment.angle = 20,
                  max.iter = 1e5)

# Saving final plot
ggsave(filename = "burke4B_25Awt_25Ako.pdf", plot = p5, path = "Plots/" )

p5
```



#################################
## GO enrichment analysis
#################################

```{r GO enrichment}
# The code below selects a list of gene IDs based on a specific Log2FC cutoff for running Overrepresentation analysis with the R function 'enricher'. The list of genes are stored in the variable "gene" for gene IDs = Ensembl gene IDs, and in the variable  'gene.entrezid' for a list of Entrez gene IDs. Usually, geneset databases use eithr Entrez gene IDs or gene symbols as identifiers, so in some cases it is handy to count with a list of genes using Entrez gene IDs. 

#The same code also extract the entire list of genes used for the Differential Expression Analysis as an R list element, using Ensemble gene IDs as keys (names) and Log2FC as values. This geneList is used as input for running Gene Set Enrichment Analysis with the function 'GSEA'. If needed, the code below also generates the same geneList but using Entrez gene IDs as keys.

go.class = list()
go.overrep = list()
go.gsea = list()

for (i in names(DE_results)){
 
  dir.create(path = paste0("./GO/PAPER/", i), showWarnings = FALSE, recursive = TRUE)
  dir.create(path = paste0("./GO/", i), showWarnings = FALSE, recursive = TRUE)
  
  for (ontho in c("BP", "MF", "CC")){ 
    my_id = paste0(i,"_",ontho)
    
    print(paste("go_classification", i, ontho))
    go.class[my_id] <- list(go_classification(dds_res = DE_results[[i]], 
                             my_file = i, 
                             onthology = ontho))
    
    print(paste("go_overrep", i, ontho))
    go.overrep[my_id] <- list(go_overrep(dds_res = DE_results[[i]], 
                                   my_file = i, 
                                   onthology = ontho))
    
    print(paste("go_gsea", i, ontho))
    go.gsea[my_id] <- list(go_gsea(dds_res = DE_results[[i]], 
                                   my_file = i, 
                                   onthology = ontho))
    
    print(paste("PAPER_plot_gsea", i, ontho))
    # Generate barplot for paper; skip ko25A_vs_koNone samples
    if(i != "ko25A_vs_koNone"){
      PAPER_plot_gsea(gsea_result = go.gsea[[my_id]], 
                      comparison_id = i, 
                      analysis_type = paste0("GO_",ontho),
                      my_x_label = paste0("Gene Ontology (", ontho, ")"),
                      my_path = "./PAPER")

      print(paste("plot_simplified_network", i, ontho))
      p.simp_net_plot <- plot_simplified_network(enrichment_result = go.gsea[[my_id]],
                                                 top_categories = 30, my_file = i)
      print(paste("ggsave2", i, ontho))
      if(typeof(p.simp_net_plot) == "list"){
        ggsave2(filename = paste0(i,"_simplified_network.pdf"), 
              plot = p.simp_net_plot, path = "./PAPER/", height = 11, width = 9)
      }
    }
    
    print(paste("Done with", my_id))
  }
}

```


I just deleted the rest of the analysis (as in RNAseL_analysis_v3.Rmd) that was not required for this ticket (TK_106)


```{r}
sessionInfo()
```